A Comparison of Regression models for Cylindrical data in Psychology.

### 2018-07-02 Manuscript/ManuscriptMarkdownDifferentSE.tex
Manuscript Update

---
title: "A Comparison of Regression models for Cylindrical data in Psychology."
author: |
  | Jolien Cremers^[Corresponding author: j.cremers@uu.nl] $^1$, Helena J.M. Pennings$^{2,3}$,  Christophe Ley $^{4}$
  | $^1$Department of Methodology and Statistics, Utrecht University
  | $^2$TNO
  | $^3$Department of Education, Utrecht University
  | $^4$Department of Applied Mathematics, Computer Science and Statistics, Ghent University

bibliography: CircularData.bib
csl: apa.csl
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
  html_document: default
  word_document: default
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{multirow}
  - \usepackage{appendix}
  - \usepackage{color}
  - \usepackage{hyperref}
  - \usepackage{subcaption}

keywords: bla vla
documentclass: article
pandoc_args: --natbib
abstract: |
  Cylindrical dats is multivariate data which consists of a directional, in this paper circular, component   and a linear component. Examples of cylindrical data in psychology include human navigation (direction     and distance of movement), eye-tracking research (direction and length of saccades) and data from an       interpersonal circumplex (type and strength of interpersonal behavior). In this paper we adapt four        models for cylindrical data to include a regression of the circular and linear component onto a set of     covariates. Subsequently, we illustrate how to fit these models and interpret their results on a dataset   on the interpersonal behavior of teachers.
---
```{r inlude = FALSE, echo = FALSE}

knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, echo = FALSE}
library(circular)
library(bpnreg)
library(haven)
library(tikzDevice)
library(plotrix)
library(tidyr)
library(dplyr) #sample_n()
library(MASS) #mvrnorm()
library(kableExtra) #latex tables
library(plotrix)
```

@@

Cylindrical data are data that consist of a linear variable and a directional 
variable. In this paper, the directional variable is circular meaning that it 
consists of a single angle instead of a set of angles. A circular variable is 
different from a linear variable in the sense that it is measured on a different
scale. Figure \ref{circline} shows the difference between a circular scale 
(right) and a linear scale (left). The most important difference is that on a 
circular scale the datapoints 0$^\circ$ and 360$^\circ$ are connected and in 
fact represent the same number while on a linear scale the two ends, $-\infty$ 
and $\infty$ are not connected. This difference requires us to use different 
statistical methods for circular variables (see e.g. @fisher1995statistical for 
an introduction to circular data and @mardia2000directional,
@jammalamadaka2001topics and @ley2017modern for a more elaborate overview).

Cylindrical data occur in several fields of research, such as for instance in 
meteorology [@garcia2013exploring], ecology [@garcia2014test] or marine research
[@lagona2015hidden]. Several types of data in psychology are also of a 
cylindrical nature. For example, in research on human navigation in the field of
cognitive psychology both distance, a linear variable, and direction, a circular
variable, of movement are of interest [@chrastil2017rotational]. In eye-tracking
research, which can be used for investigating various cognitive processes (e.g. 
those involved in reading a text), we can also speak of cylindrical data. When 
measuring eye-movements we speak of the eye-movements themselves, the saccades, 
and fixations, the periods of time between movements when the eyes are looking 
at one point. Of the saccades both the direction, a circular variable, and the 
duration, a linear variable, are of interest (for a review of eye-tracking 
research see @rayner200935th). Data from circumplex measurement instruments,
e.g. the interpersonal circumplex as used in personality psychology, are also of
a cylindrical nature (see Section \ref{Example} for a more detailed
explanation).

@@

As is the case for circular data, the analysis of cylindrical data requires 
special methods. Several methods have been put forward to model the relation 
between the linear and circular component of a cylindrical variable. Some of 
these are based on regressing the linear component onto the circular component 
using the following type of relation: $$y = \beta_0 + \beta_1*\cos(\theta) + 
\beta_2*\sin(\theta)+ \epsilon,$$ where $y$ is the linear component and $\theta$
the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a 
different way, e.g. by specifying a multivariate model for several linear and 
circular variables and modelling their covariance matrix 
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. For 
example, @abe2017tractable introduce a cylindrical distribution based on a 
Weibull distribution for the linear component and a sine-skewed von Mises 
distribution for the circular component and link these through their shape and 
concentration parameters respectively. However, none of the methods that have 
been proposed thus far include additional covariates onto which both the
circular and linear component are regressed.

@@

The motivating example for this article comes from the field of educational 
psychology and was collected for the studies on classroom climate of
@pennings2017interpersonal, @Claessens2016side and @vanderWant2015role. An 
indicator of classroom climate is the students' perceptions of their teachers'
interpersonal behavior. These interpersonal perceptions, both in educational
psychology as well as in other areas of psychology, can be measured using
circumplex measurement instruments (see
@horowitz2010handbook for an overview of many such instruments).

The Questionnaire on Teacher Interaction (QTI) [@wubbels2006interpersonal] is
one such circumplex measurement instrument. It is used to study student
perceptions of their teachers' interpersonal behavior and contains
items that load on two dimensions: Agency and Communion. Agency refers to the
degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two 
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. Different parts of
this space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. We call the two-dimensional space the interpersonal circle (IPC).
The idea is that the IPC is ``a continuous order with no beginning or end''
[@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering and
the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular variable.

@@

@Cremers2018Assessing show how data from the IPC can be considered circular data
and analyzed as such using a regression model. The two-dimension scores Agency
and Communion can be converted to a circular score using the two-argument
arctangent function in \eqref{PredVal}, where $A$ represents a score on
the Agency dimension and $C$ represents a score on the Communion dimension. The
resulting circular variable $\theta$ can then be modeled. However, when
two-dimensional data are converted to the circle we lose some information, the
length of the two-dimensional vector $(A, C)$, \emph{i.e.}, its euclidean norm
$\mid\mid (A, C) \mid\mid$. This length represents the strength of the type of
interpersonal behavior a teacher shows towards his/her students. In a
cylindrical model we are able to incorporate this information, and model a
circular variable $\theta$ together with a linear variable corresponding to
$\mid\mid (A, C) \mid\mid$. This leads to an improved analysis of data from the
IPC. In the next section we introduce several models that can be used for more
accurate and informative regression analysis on the teacher data. Descriptives
for the teacher data are given in Section \ref{DataAnalysis}.

@@

```{r loaddata, return = FALSE, echo = FALSE}
Dat <- read_spss("Data Heleen/mergeddata.sav")

#only select the first measurement occasion
Dat <- subset(Dat, Time == 0)

#sort on class size
Dat <- Dat[order(-Dat$nklas), ]

#Now remove duplicates 
#(because dataframe is sorted on class size the smallest classes are removed)

duplicates <- duplicated(Dat[,c("docnr", "Time")])
Dat <- Dat[!duplicates,]

#center and remove missings for self-efficacy
Dat <- Dat %>% drop_na(Efficacy_CM)
Dat$SEc <- Dat$Efficacy_CM- mean(Dat$Efficacy_CM)

set.seed(50)

#Create holdout and test sample
Dat.hold <- sample_n(Dat, 15)
Dat.test <- Dat[-as.numeric(rownames(Dat.hold)),]

#Compute circular and linear outcome
theta.test <- atan2(Dat.test$lAgency, Dat.test$lcommunion)
y.test     <- sqrt(Dat.test$lAgency^2 + Dat.test$lcommunion^2)

theta.hold <- atan2(Dat.hold$lAgency, Dat.hold$lcommunion)
y.hold     <- sqrt(Dat.hold$lAgency^2 + Dat.hold$lcommunion^2)

```

@@

In this section we introduce four cylindrical regression models to contain 
predictors for the linear and circular outcomes, $\Theta$ and $Y$. We extend the models from @mastrantonio2018joint and
@abe2017tractable. Additionally, we introduce two models where the relation
between $\Theta$ and $Y$ is modelled as follows (following
@mastrantonio2015bayesian):

@@

In both of these models the relation between $\Theta \in [0, 2\pi)$ and $Y\in 
(-\infty, + \infty)$ is specified as in \eqref{circlinlink} with the following
conditional distribution:

@@

In a regression setup the outcome $\theta_i$ for each individual $i = 1, \dots,
n$, where $n$ is the sample size, is generated independently from \eqref{pnreg}.
The mean vector $\boldsymbol{\mu}_i \in \mathbb{R}^2$ is defined as
$\boldsymbol{\mu}_i = \boldsymbol{z}_i\boldsymbol{B}$. The vector 
$\boldsymbol{z}_i$ is a a vector of $p$ covariate values and $\boldsymbol{B} = 
(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})$ contain the regression 
coefficients. Note however that the dimensions of $\boldsymbol{\beta}^{I}$ and 
$\boldsymbol{\beta }^{I}$ need not necessarily be the same and we are thus 
allowed to have a different set of predictor variables and vectors 
$\boldsymbol{z}_i^I$ and $\boldsymbol{z}_i^{II}$ for the two components of 
$\boldsymbol{\mu}_i$.

Following @wang2012directional and @hernandez2016general the joint density of
$r$ and $\theta$ for the GPN distribution equals:

@@

where $\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2  & \rho\\ \rho & 1 
\end{bmatrix}$, $\boldsymbol{u}= \begin{bmatrix} \cos \theta \\ \sin \theta 
\end{bmatrix}$. In a regression setup the outcome $\theta_i$ for each individual
is generated independently from \eqref{gpnreg}. The mean vector 
$\boldsymbol{\mu}_i \in \mathbb{R}^2$ is defined as $\boldsymbol{\mu}_i = 
\boldsymbol{z}_i(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})$. The vector 
$\boldsymbol{z}_i$ is a a vector of $p$ covariate values and each 
$\boldsymbol{\beta}^{k}$ is a vector with two regression coefficients, one for 
each of the two components of $\boldsymbol{\mu}_i$. Note that for the CL-GPN 
model we do need to have the same predictors for both components of 
$\boldsymbol{\mu}_i$.

Both cylindrical models introduced here are estimated using MCMC methods based
on @nunez2011bayesian, @wang2012directional and @hernandez2016general for the 
regression of the circular outcome. A detailed description of the Bayesian
estimation and MCMC samplers can be found in Appendices \ref{A1} and \ref{A2}.

@@

This model is an extension of the cylindrical model introduced by
@abe2017tractable to the regression context. The joint density of $\Theta$ and
$Y$, in this model defined on the positive real half-line $[0, + \infty)$, is:

@@

In a regression setup the outcome vector $(\theta_i, y_i)^t$ for each individual
is generated independently from \eqref{WeiSSVMdensity}. The parameters $\alpha >
0$ and $\beta_i = \exp(\boldsymbol{x}_i^t\boldsymbol{\nu}) > 0$ are linear shape
and scale parameters, $\mu_i = \eta_0 + 
2\tan^{-1}(\boldsymbol{z}_i^t\boldsymbol{\eta}) \in [0, 2\pi)$, $\kappa > 0$ and
$\lambda \in [-1, 1]$ are circular location, concentration and skewness 
parameters. The parameter $\boldsymbol{\nu}$ is a vector of $q$ regression 
coefficients $\nu_j \in (-\infty, +\infty)$ for the prediction of $y$ where $j =
0, \dots, q$ and $\nu_0$ is the intercept. The parameter $\eta_0 \in [0, 2\pi)$
is the intercept and $\boldsymbol{\eta}$ is a vector of $p$ regression
coefficients $\eta_j \in (-\infty, +\infty)$ for the prediction of $\theta$
where $j = 1, \dots, p$. The vector $\boldsymbol{x}_i$ is a vector of predictor
values for the prediction of $y$ and $\boldsymbol{z}_i$ is a vector of predictor
values for the prediction of $\theta$.

As in @abe2017tractable, the conditional distribution of $y$ given $\theta$ is a
Weibull distribution with scale $\alpha$ and shape
$\beta(1-\tanh(\kappa)\cos(\theta - \mu))^{1-\alpha}$ and the conditional 
distribution of $\theta$ given $y$ is a sine skewed von Mises distribution with 
location parameter $\mu$ and concentration parameter $(\beta 
y)^\alpha\tanh(\kappa)$.

@@

This model is an extension of the cylindrical model introduced by
@mastrantonio2018joint to the regression context. It contains $p$ circular
outcomes and $q$ linear outcomes. The circular outcomes $\boldsymbol{\Theta} = 
(\boldsymbol{\Theta}_1, \dots, \boldsymbol{\Theta}_p)$ are modeled together by a
multivariate GPN distribution and the linear outcomes $\boldsymbol{Y} = 
(\boldsymbol{Y}_1, \dots, \boldsymbol{Y}_q)$  are modeled together by a 
multivariate skew normal distribution [@sahu2003new]. Because the GPN 
distribution is modelled using a so-called augmented representation (as in 
\eqref{projection} and \eqref{gpnreg}) it is convenient to use a similar tactic
for modelling the multivariate skew normal distribution. Following
@mastrantonio2018joint the linear outcomes are represented as:

$$\boldsymbol{Y} = \boldsymbol{\mu}_y + \boldsymbol{\Lambda}\boldsymbol{D} + \boldsymbol{H},$$
where where $\boldsymbol{\mu}_y$ is a mean vector for the linear outcome 
$\boldsymbol{Y}$, $\boldsymbol{\Lambda} = \text{diag}(\boldsymbol{\lambda})$ is a $q 
\times q$ diagonal matrix with diagonal elements $\boldsymbol{\lambda} = 
\lambda_1, \dots, \lambda_q$ (skewness parameters), $\boldsymbol{D} \sim HN_q(\boldsymbol{0}_q, 
\boldsymbol{I}_q)$, a q-dimensional half normal distribution
[@olmos2012extension], and $\boldsymbol{H} \sim N_q(\boldsymbol{0}_q,
\boldsymbol{\Sigma}_y)$. This means that conditional on the auxiliary data
$\boldsymbol{D}$, $\boldsymbol{Y}$ is normally distributed with mean
$\boldsymbol{\mu}_y + \boldsymbol{\Lambda}\boldsymbol{D}$ and covariance matrix
$\boldsymbol{\Sigma}_y$. The joint density for $(\boldsymbol{Y},
\boldsymbol{D})^t$ is defined as follows:

@@

As in @mastrantonio2018joint dependence between the linear and circular outcome
is created by modelling the augmented representations of $\boldsymbol{\Theta}$,
and $\boldsymbol{Y}$ together in a $2 \times p + q$ dimensional normal
distribution. The joint density of the model is then represented by:

@@

where the mean vector $\boldsymbol{\mu} = (\boldsymbol{\mu}_s,
\boldsymbol{\mu}_y)^t$ and $\boldsymbol{\Sigma} = \left ( \begin{matrix}
\boldsymbol{\Sigma}_s & \boldsymbol{\Sigma}_{sy} \\ \boldsymbol{\Sigma}_{sy^t} &
\boldsymbol{\Sigma}_y \\ \end{matrix} \right )$. The matrix
$\boldsymbol{\Sigma}_s$ is the covariance matrix for the variances of and
covariances between the augmented representations of the circular outcome and
the matrix $\boldsymbol{\Sigma}_{sy}$ contains covariances between the augmented
representations of the circular outcome and the linear outcome.

In our regression extension we have $i = 1, \dots, n$ observations of $p$ 
circular outcomes, $q$ linear outcomes and $k$ covariates. The mean in the 
density in \eqref{YDThetarjoint} then becomes $\boldsymbol{\mu}_i = 
\boldsymbol{x}_i\boldsymbol{B}$ where $\boldsymbol{B}$ is a $k \times (2 
\times p + q)$ matrix with regression coeffients. We estimate the model using 
MCMC methods. A detailed description of these methods is given in Appendix
\ref{A3}.

@@

where $l$ is the model likelihood, $M$ is the sample size of the holdout set, 
$x_i$ is the $i^{th}$ datapoint from the holdout set and 
$\hat{\boldsymbol{\vartheta}}$ are the ML estimates of the model parameters. 
Using posterior samples the criterion is similar to the log pointwise predictive
density (lppd) as outlined in @BDA and can be computed as:

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={A Comparison of Regression models for Cylindrical data in Psychology.},
            pdfkeywords={bla vla},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{A Comparison of Regression models for Cylindrical data in Psychology.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Jolien Cremers\footnote{Corresponding author:
  \href{mailto:j.cremers@uu.nl}{\nolinkurl{j.cremers@uu.nl}}} \(^1\),
Helena J.M. Pennings\(^{2,3}\), Christophe Ley \(^{4}\)\\
\(^1\)Department of Methodology and Statistics, Utrecht University\\
\(^2\)TNO\\
\(^3\)Department of Education, Utrecht University\\
\(^4\)Department of Applied Mathematics, Computer Science and
Statistics, Ghent University}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\usepackage{multirow}
\usepackage{appendix}
\usepackage{color}
\usepackage{hyperref}
\usepackage{subcaption}

\begin{document}
\maketitle
\begin{abstract}
Cylindrical dats is multivariate data which consists of a directional,
in this paper circular, component and a linear component. Examples of
cylindrical data in psychology include human navigation (direction and
distance of movement), eye-tracking research (direction and length of
saccades) and data from an interpersonal circumplex (type and strength
of interpersonal behavior). In this paper we adapt four models for
cylindrical data to include a regression of the circular and linear
component onto a set of covariates. Subsequently, we illustrate how to
fit these models and interpret their results on a dataset on the
interpersonal behavior of teachers.
\end{abstract}

@@

Cylindrical data are data that consist of a linear variable and a
directional variable. In this paper, the directional variable is
circular meaning that it consists of a single angle instead of a set of
angles. A circular variable is different from a linear variable in the
sense that it is measured on a different scale. Figure \ref{circline}
shows the difference between a circular scale (right) and a linear scale
(left). The most important difference is that on a circular scale the
datapoints 0\(^\circ\) and 360\(^\circ\) are connected and in fact
represent the same number while on a linear scale the two ends,
\(-\infty\) and \(\infty\) are not connected. This difference requires
us to use different statistical methods for circular variables (see e.g.
Fisher (1995) for an introduction to circular data and Mardia \& Jupp
(2000), Jammalamadaka \& Sengupta (2001) and Ley \& Verdebout (2017) for
a more elaborate overview).

Cylindrical data occur in several fields of research, such as for
instance in meteorology (García-Portugués, Crujeiras, \&
González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras,
González-Manteiga, \& Pereira, 2014) or marine research (Lagona, Picone,
Maruotti, \& Cosoli, 2015). Several types of data in psychology are also
of a cylindrical nature. For example, in research on human navigation in
the field of cognitive psychology both distance, a linear variable, and
direction, a circular variable, of movement are of interest (Chrastil \&
Warren, 2017). In eye-tracking research, which can be used for
investigating various cognitive processes (e.g. those involved in
reading a text), we can also speak of cylindrical data. When measuring
eye-movements we speak of the eye-movements themselves, the saccades,
and fixations, the periods of time between movements when the eyes are
looking at one point. Of the saccades both the direction, a circular
variable, and the duration, a linear variable, are of interest (for a
review of eye-tracking research see Rayner (2009)). Data from circumplex
measurement instruments, e.g.~the interpersonal circumplex as used in
personality psychology, are also of a cylindrical nature (see Section
\ref{Example} for a more detailed explanation).

@@

As is the case for circular data, the analysis of cylindrical data
requires special methods. Several methods have been put forward to model
the relation between the linear and circular component of a cylindrical
variable. Some of these are based on regressing the linear component
onto the circular component using the following type of relation:
\[y = \beta_0 + \beta_1*\cos(\theta) + 
\beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is the linear component
and \(\theta\) the circular component (Johnson \& Wehrly, 1978; Mardia
\& Sutton, 1978; Mastrantonio, Maruotti, \& Jona-Lasinio, 2015). Others
model the relation in a different way, e.g.~by specifying a multivariate
model for several linear and circular variables and modelling their
covariance matrix (Mastrantonio, 2018) or by proposing a joint
cylindrical distribution. For example, Abe \& Ley (2017) introduce a
cylindrical distribution based on a Weibull distribution for the linear
component and a sine-skewed von Mises distribution for the circular
component and link these through their shape and concentration
parameters respectively. However, none of the methods that have been
proposed thus far include additional covariates onto which both the
circular and linear component are regressed.

@@

The motivating example for this article comes from the field of
educational psychology and was collected for the studies on classroom
climate of Pennings et al. (2017), Claessens (2016) and Want (2015). An
indicator of classroom climate is the students' perceptions of their
teachers' interpersonal behavior. These interpersonal perceptions, both
in educational psychology as well as in other areas of psychology, can
be measured using circumplex measurement instruments (see Horowitz \&
Strack (2011) for an overview of many such instruments).

The Questionnaire on Teacher Interaction (QTI) (Wubbels, Brekelmans,
Brok, \& Tartwijk, 2006) is one such circumplex measurement instrument.
It is used to study student perceptions of their teachers' interpersonal
behavior and contains items that load on two dimensions: Agency and
Communion. Agency refers to the degree of power or control a teacher
exerts in interaction with his/her students. Communion refers to the
degree of friendliness or affiliation a teacher conveys in interaction
with his/her students. The loadings on the two dimensions of the QTI can
be placed in a two-dimensional space formed by Agency (vertical) and
Communion (horizontal), see Figure \ref{QTI}. Different parts of this
space are characterized by different teacher behavior, e.g. `helpful' or
`uncertain'. We call the two-dimensional space the interpersonal circle
(IPC). The idea is that the IPC is ``a continuous order with no
beginning or end'' (Gurtman, 2009, p. 2). We call such ordering a
circumplex ordering and the IPC is therefore often called the
interpersonal circumplex. The ordering also implies that scores on the
IPC could be viewed as a circular variable.

@@

Cremers, Mainhard, \& Klugkist (2018a) show how data from the IPC can be
considered circular data and analyzed as such using a regression model.
The two-dimension scores Agency and Communion can be converted to a
circular score using the two-argument arctangent function in
\eqref{PredVal}, where \(A\) represents a score on the Agency dimension
and \(C\) represents a score on the Communion dimension. The resulting
circular variable \(\theta\) can then be modeled. However, when
two-dimensional data are converted to the circle we lose some
information, the length of the two-dimensional vector \((A, C)\),
\emph{i.e.}, its euclidean norm \(\mid\mid (A, C) \mid\mid\). This
length represents the strength of the type of interpersonal behavior a
teacher shows towards his/her students. In a cylindrical model we are
able to incorporate this information, and model a circular variable
\(\theta\) together with a linear variable corresponding to
\(\mid\mid (A, C) \mid\mid\). This leads to an improved analysis of data
from the IPC. In the next section we introduce several models that can
be used for more accurate and informative regression analysis on the
teacher data. Descriptives for the teacher data are given in Section
\ref{DataAnalysis}.

@@

@@

In this section we introduce four cylindrical regression models to
contain predictors for the linear and circular outcomes, \(\Theta\) and
\(Y\). We extend the models from Mastrantonio (2018) and Abe \& Ley
(2017). Additionally, we introduce two models where the relation between
\(\Theta\) and \(Y\) is modelled as follows (following Mastrantonio et
al. (2015)):

@@

In both of these models the relation between \(\Theta \in [0, 2\pi)\)
and \(Y\in (-\infty, + \infty)\) is specified as in \eqref{circlinlink}
with the following conditional distribution:

@@

In a regression setup the outcome \(\theta_i\) for each individual
\(i = 1, \dots, n\), where \(n\) is the sample size, is generated
independently from \eqref{pnreg}. The mean vector
\(\boldsymbol{\mu}_i \in \mathbb{R}^2\) is defined as
\(\boldsymbol{\mu}_i = \boldsymbol{z}_i\boldsymbol{B}\). The vector
\(\boldsymbol{z}_i\) is a a vector of \(p\) covariate values and
\(\boldsymbol{B} = (\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})\)
contain the regression coefficients. Note however that the dimensions of
\(\boldsymbol{\beta}^{I}\) and \(\boldsymbol{\beta }^{I}\) need not
necessarily be the same and we are thus allowed to have a different set
of predictor variables and vectors \(\boldsymbol{z}_i^I\) and
\(\boldsymbol{z}_i^{II}\) for the two components of
\(\boldsymbol{\mu}_i\).

Following Wang \& Gelfand (2013) and Hernandez-Stumpfhauser et al.
(2016) the joint density of \(r\) and \(\theta\) for the GPN
distribution equals:

@@

where
\(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2 & \rho\\ \rho & 1 \end{bmatrix}\),
\(\boldsymbol{u}= \begin{bmatrix} \cos \theta \\ \sin \theta \end{bmatrix}\).
In a regression setup the outcome \(\theta_i\) for each individual is
generated independently from \eqref{gpnreg}. The mean vector
\(\boldsymbol{\mu}_i \in \mathbb{R}^2\) is defined as
\(\boldsymbol{\mu}_i = \boldsymbol{z}_i(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})\).
The vector \(\boldsymbol{z}_i\) is a a vector of \(p\) covariate values
and each \(\boldsymbol{\beta}^{k}\) is a vector with two regression
coefficients, one for each of the two components of
\(\boldsymbol{\mu}_i\). Note that for the CL-GPN model we do need to
have the same predictors for both components of \(\boldsymbol{\mu}_i\).

Both cylindrical models introduced here are estimated using MCMC methods
based on Nuñez-Antonio et al. (2011), Wang \& Gelfand (2013) and
Hernandez-Stumpfhauser et al. (2016) for the regression of the circular
outcome. A detailed description of the Bayesian estimation and MCMC
samplers can be found in Appendices \ref{A1} and \ref{A2}.

@@

This model is an extension of the cylindrical model introduced by Abe \&
Ley (2017) to the regression context. The joint density of \(\Theta\)
and \(Y\), in this model defined on the positive real half-line
\([0, + \infty)\), is:

@@

In a regression setup the outcome vector \((\theta_i, y_i)^t\) for each
individual is generated independently from \eqref{WeiSSVMdensity}. The
parameters \(\alpha > 0\) and
\(\beta_i = \exp(\boldsymbol{x}_i^t\boldsymbol{\nu}) > 0\) are linear
shape and scale parameters,
\(\mu_i = \eta_0 + 2\tan^{-1}(\boldsymbol{z}_i^t\boldsymbol{\eta}) \in [0, 2\pi)\),
\(\kappa > 0\) and \(\lambda \in [-1, 1]\) are circular location,
concentration and skewness parameters. The parameter
\(\boldsymbol{\nu}\) is a vector of \(q\) regression coefficients
\(\nu_j \in (-\infty, +\infty)\) for the prediction of \(y\) where
\(j = 0, \dots, q\) and \(\nu_0\) is the intercept. The parameter
\(\eta_0 \in [0, 2\pi)\) is the intercept and \(\boldsymbol{\eta}\) is a
vector of \(p\) regression coefficients
\(\eta_j \in (-\infty, +\infty)\) for the prediction of \(\theta\) where
\(j = 1, \dots, p\). The vector \(\boldsymbol{x}_i\) is a vector of
predictor values for the prediction of \(y\) and \(\boldsymbol{z}_i\) is
a vector of predictor values for the prediction of \(\theta\).

As in Abe \& Ley (2017), the conditional distribution of \(y\) given
\(\theta\) is a Weibull distribution with scale \(\alpha\) and shape
\(\beta(1-\tanh(\kappa)\cos(\theta - \mu))^{1-\alpha}\) and the
conditional distribution of \(\theta\) given \(y\) is a sine skewed von
Mises distribution with location parameter \(\mu\) and concentration
parameter \((\beta y)^\alpha\tanh(\kappa)\).

@@

This model is an extension of the cylindrical model introduced by
Mastrantonio (2018) to the regression context. It contains \(p\)
circular outcomes and \(q\) linear outcomes. The circular outcomes
\(\boldsymbol{\Theta} = (\boldsymbol{\Theta}_1, \dots, \boldsymbol{\Theta}_p)\)
are modeled together by a multivariate GPN distribution and the linear
outcomes
\(\boldsymbol{Y} = (\boldsymbol{Y}_1, \dots, \boldsymbol{Y}_q)\) are
modeled together by a multivariate skew normal distribution (Sahu, Dey,
\& Branco, 2003). Because the GPN distribution is modelled using a
so-called augmented representation (as in \eqref{projection} and
\eqref{gpnreg}) it is convenient to use a similar tactic for modelling
the multivariate skew normal distribution. Following Mastrantonio (2018)
the linear outcomes are represented as:

\[\boldsymbol{Y} = \boldsymbol{\mu}_y + \boldsymbol{\Lambda}\boldsymbol{D} + \boldsymbol{H},\]
where where \(\boldsymbol{\mu}_y\) is a mean vector for the linear
outcome \(\boldsymbol{Y}\),
\(\boldsymbol{\Lambda} = \text{diag}(\boldsymbol{\lambda})\) is a
\(q \times q\) diagonal matrix with diagonal elements
\(\boldsymbol{\lambda} = \lambda_1, \dots, \lambda_q\) (skewness
parameters),
\(\boldsymbol{D} \sim HN_q(\boldsymbol{0}_q, \boldsymbol{I}_q)\), a
q-dimensional half normal distribution (Olmos, Varela, Gómez, \&
Bolfarine, 2012), and
\(\boldsymbol{H} \sim N_q(\boldsymbol{0}_q, \boldsymbol{\Sigma}_y)\).
This means that conditional on the auxiliary data \(\boldsymbol{D}\),
\(\boldsymbol{Y}\) is normally distributed with mean
\(\boldsymbol{\mu}_y + \boldsymbol{\Lambda}\boldsymbol{D}\) and
covariance matrix \(\boldsymbol{\Sigma}_y\). The joint density for
\((\boldsymbol{Y}, \boldsymbol{D})^t\) is defined as follows:

@@

As in Mastrantonio (2018) dependence between the linear and circular
outcome is created by modelling the augmented representations of
\(\boldsymbol{\Theta}\), and \(\boldsymbol{Y}\) together in a
\(2 \times p + q\) dimensional normal distribution. The joint density of
the model is then represented by:

@@

where the mean vector
\(\boldsymbol{\mu} = (\boldsymbol{\mu}_s, \boldsymbol{\mu}_y)^t\) and
\(\boldsymbol{\Sigma} = \left ( \begin{matrix} \boldsymbol{\Sigma}_s & \boldsymbol{\Sigma}_{sy} \\ \boldsymbol{\Sigma}_{sy^t} & \boldsymbol{\Sigma}_y \\ \end{matrix} \right )\).
The matrix \(\boldsymbol{\Sigma}_s\) is the covariance matrix for the
variances of and covariances between the augmented representations of
the circular outcome and the matrix \(\boldsymbol{\Sigma}_{sy}\)
contains covariances between the augmented representations of the
circular outcome and the linear outcome.

In our regression extension we have \(i = 1, \dots, n\) observations of
\(p\) circular outcomes, \(q\) linear outcomes and \(k\) covariates. The
mean in the density in \eqref{YDThetarjoint} then becomes
\(\boldsymbol{\mu}_i = \boldsymbol{x}_i\boldsymbol{B}\) where
\(\boldsymbol{B}\) is a \(k \times (2 \times p + q)\) matrix with
regression coeffients. We estimate the model using MCMC methods. A
detailed description of these methods is given in Appendix \ref{A3}.

@@

where \(l\) is the model likelihood, \(M\) is the sample size of the
holdout set, \(x_i\) is the \(i^{th}\) datapoint from the holdout set
and \(\hat{\boldsymbol{\vartheta}}\) are the ML estimates of the model
parameters. Using posterior samples the criterion is similar to the log
pointwise predictive density (lppd) as outlined in Gelman et al. (2014)
and can be computed as:
